---
title: "3: Tabular Data"
format: html
editor: visual
---

So far we have used individual vectors as data, but of course it is not particularly convenient to work with real data like that. Whether your data originally is stored in an Excel file or some sort of database system, it is usually stored in a *tabular* format, ideally with two key properties:

1.  Each **column** of your data is a **variable** (like age, weight, ...)
2.  Each **row** of your data is an **observation** (like a single patient, a hospital, a country, ...)

In the real world, data tends to come in all shapes and sizes, but for the purposes of analysis this is the shape to strive for.

Just like we made up some patient data to store as vectors, we can use the same approach to create tabular data, using the `data.frame()` function. Each argument of this function takes a *vector* just like we used before, so in a sense the `data.frame` is just a way to organize vectors that all must have the same length!

```{r df-ex1}
example_df <- data.frame(
  id = 1:5,
  age = c(35, 32, 26, 40, 34),
  weight_kg = c(75, 64, 71, 45, 85)
)

example_df
```

## Working with Tabular Data

We manipulated vectors by indexing them with `[ ]`, things are more complicated with `data.frames` but fundamentally the idea is the same!

We can access individual **columns** of the data.frame like the vectors they are using either `$` or *double* brackets for indexing `[[ ]]` where the index can be the integer index or *the name* of the variable, so these are all identical and result in a vector:

```{r df-access-columns}
example_df$age
example_df[[2]]
example_df[["age"]]
```

Since this gives us a vector, we can do the usual vector things:

```{r df-vector-subset}
# First 5 age values
example_df$age[1:5]

# All age values under 30
example_df$age[example_df$age < 30]

# 2nd column, 3rd element
example_df[[2]][3]
```

This becomes inconvenient really quickly though, which is why there are better ways we'll see soon.

Note that this way we can only ever get *one* vector, so this does **not work** as we might expect:

```{r, eval=FALSE}
example_df[[c(1, 2)]]
example_df[[c("id", "age")]]
```

We can however use single brackets to access one *or more* columns, rows, or both, and it get's a little more complex:

```{r df-access-colrow}
# First column, result is a data.frame
example_df[1]
example_df["id"]
# 3rd row
example_df[3, ]
# First column, 5th row
example_df[5, 1]
example_df[5, "id"]
```

When we use single brackets, the result is still a `data.frame` unless in the last case where the result is a scalar. This can become somewhat tricky, but if you *insist* on the result being a `data.frame` you can add `drop = FALSE`:

```{r df-drop}
example_df[5, 1, drop = FALSE]
```

This is a "good to know and easy to forget" bit of information.
Soon we will introduce mechanisms to make working with tabular data a bit more intuitive, but since you will most likely encounter this type of code in the wild, we'll practice it a little soon.

## External Data

### Built-in

R comes with some built-in datasets that are "just there":

```{r df-builtin}
sleep
mtcars
iris
```

They are built-in in the sense that they are part of the R installation - you can not accidentally delete them and you don't need to do anything special to access them, which makes them the standard case for tutorials and examples --- and of course also a little boring!

We can also install packages that provide additional data, and many packages that provide some sort of statistical or visualization functionality often include toy datasets to illusrate their use.  
We'll see examples of that later as well!

### Reading and Writing Data from/to Files

A usual part of the workflow is to import data, clean it up, modify it in some way, and then storing it as a file for later use or to share it with collaborators.

There are many formats we could use, and they vary in how simple they are to use within R, how easy it is to read them with other software, and how certain you want to be that people are going to be able to read the data in 30 years.

For simplicity we'll focus on two kinds of formats: R's built-in format `RDS`, and plain-text delimited formats like CSV and TSV.

#### RDS

RDS has the benefit of being able to losslessly store any kidn of R objects, which means that whatever you store as a file on your computer you'll most likely be able to just read into R again and use it as if nothing happened.  
This is why RDS is the most popular for "within R" workflow where you want to store intermediate results for example, or cleaned datasets you want to use later for analysis in a different R project.

We can store and restore data to RDS like this, using the `example_df` and the `data` directory in this project:

```{r rds-store}
saveRDS(example_df, file = "data/example.rds")
```

```{r rds-read}
example_rds <- readRDS(file = "data/example.rds")
```

We can check if the two are identical:

```{r}
identical(example_df, example_rds)
```

### CSV

Comma separated values (CSV) are maybe the most common way to share data that's not large enough to require database or strong compression.
If you want to share data in a way that others can import it into Excel without a big headache, this is probably the way to go!  
Also if you want to make sure that your great-great-great-grandchildren are still able to read your data without any special software.

It's the lowest common denominator in many ways, and it can be very annoying to deal with.

In the simplest case we can read and write CSVs with `read.csv()`

```{r csv-write}
write.csv(example_df, file = "data/example.csv")
```

```{r csv-read}
example_csv <- read.csv(file = "data/example.csv")
```

Note that `?read.csv` and `?write.csv` show a few option that will be useful at some point, like `sep`.
By default, we write and read comma-delimited files, but we could also do `sep = ";"` to use `;` as a separating character!  
When you try to read data with Excel for example, then using `;` might make things easier.


When it comes to reading external data, tons of things can go wrong and for now, we just assume that it went okay.

For future reference though:

-   When in doubt, google around for how to best read the given data into R
-   Ensure to glance at the relevant documentation, e.g. for `?read.csv`, if you run into errors or unexpected behavior
-   Once you're used to installing packages, the `readr` package might be of interest here
-   Refer to [this chapter in R4DS](https://r4ds.hadley.nz/import) for more information and examples!


::: {.callout-tip}
To keep a project organized, it's good practice to 

- Store original, raw data in a dedicated directory like `data-raw` and **never modify it**!
- Read it, clean it, and store a clean version of your data in `data`

That way, you should always your original data, your final data, and most importantly: Reproducible R code you can share with others that explains how to get from point A to point B!
:::

## Excursion: Packages

Before we move on, it's time to take a look at this open source ecosystem we mentioned in the beginning. In R, this mostly means we can install *packages*. A package is a collection of functions and sometimes other objects which we can use for additional capabilities, and lets us benefit from other people's work such that we don't have to reinvent the wheel every time we do an analysis.

We can install packages using either RStudio's "Packages" pane to the right, or we can use R functions directly.

For example, there is a package called `"readxl"` which allows us to Excel files (`.xlsx`), and we can install it like this:

```{r install-package, eval=FALSE}
install.packages("readxl")
```

We can then **load** the package using the `library()` function.

```{r load-package}
library(readxl)
```

We can now use functions from the package, and if we don't know where to start, usually `?readxl` is a good entrypoint.

Keep in mind:

1.  You have to **install** a package *once* into your "library"
2.  You have to **load** the package every time you restart R
3.  You **should** occasionally *update* your installed packages using `update.packages()`.

As an example for what we would use `readxl` for, let's read an example dataset stored on `data-raw` in this folder in the `.xlsx` format.

```{r read-xlsx}
example_df <- read_excel("data-raw/example.xlsx")
```

Here, the function `read_excel()` is provided by the package `readxl` and only available when that package is installed and loaded!
An error message like

```
could not find function "read_excel"
```

indicated that you either have not loaded the required package or maybe mistyped the function name.


::: {.callout-note} 

Once a package is installed, it is also possible to use one of its functions without loading it beforehand by using the `::` syntax, like `readxl::read_excel()`

We won't do that here yet, but in the wild you might find people referring to a function in the format `somePackage::some_fun()` to clarify they mean `some_fun()` from the package `somePackage`!

:::

### Your Turn

We will now move on to using more packages, which of course first need installation.
Unless you have already done so, install the following packages:

-   "tidyverse"
-   "gapminder"
-   "gtsummary"
-   "ggstatsplot"

Optional:
-   "ggthemes"
-   "readr"

**Alternatively** you can install the package `pak` first, and then install the packages above using `pkg_install()`:

```{r install-pak, eval=FALSE}
pkg_install("tidyverse")
```

`pak` should be a bit faster and depending on your operating system also more robust, and it brings a few features not relevant yet.  
If you use a Linux distribution like Ubuntu, `pak` can save you a lot of annoyances.

In any case, this might take a moment!

## Knowing Your Data

As mentioned, accessing tabular data is a bit more complex than it is for vectors, and in the next section we will learn about convenient tools for this case.
However, there are some common built-in tools that R provides for this which you should be at least aware of.

We'll start with a new dataset from one of the packages you just installed: `gapminder`
This dataset is based on the [Gapminder](https://www.gapminder.org/) data, and also part of (Hans Rosling's almost 20 year old TED talk](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen).

First we _load the data_ which in this case means to just load the `gapminder` package to access the dataset with the same name (confusingly so) 


```{r}
library(gapminder)

gapminder
```

To see the number of rows and colums, we can use `nrow()`, `ncol()`, or `dim()`.

```{r}
nrow(gapminder)
ncol(gapminder)
dim(gapminder)
```

We often want to see the first or last few rows using `head()` and `tail()` (note: these also work on vectors!)

```{r}
head(gapminder)
tail(gapminder)
```

We can also use `str()` to get a basic overview of the data, including each variables type:

```{r}
str(gapminder)
```

Don't forget we can also refer to the documentation using `?gapminder` to learn about what each variable means!

There is also something new there: A type of data we have not encountered before: The `factor`!
`factors` are for _categorical data_, particularly the kind where we know beforehand how many different categories there are and what they are called.

A quick example:

```{r factor-example-1}
fruit <- factor(c("apple", "orange", "apple"))
fruit

levels(fruit)
```

We can set the `levels` of a factor manually or change them later.  
A `factor` can have a level that does not appear on the data, but it will "know" that there is an extra category, which helps us keep an overvoew on what we expect!

```{r factor-example-2}
fruit <- factor(
   c("orange", "apple", "apple", "orange", "apple", "orange"), 
   levels = c("apple", "banana", "orange")
)
fruit

levels(fruit)
```

The `table()` function gives us a simple frequency table of the data, and since it expects there to be `"banana"`, it shows us a frequency of 0:

```{r}
table(fruit)
```

`factors` are like `characters` but with _additional information_ stored in them, and dealing with them can be a bit counter-intuitive.
For now, we're happy with knowing that they exists and that they are useful for categorical data!

In `gapminder`, there are two `factor` variables as we have seen using `str()`: `country` and `continent`.

We can use the aforementioned `table()` to get a broad overview of what we are dealing with

```{r}
table(gapminder$continent)
```

We will get a better overview later, but for now we'll practice the basics for a moment

### Your Turn

Using the `gapminder` data, try to find out the following:

1. Select the variables `country`, `pop` and `lifeExp`
2. Filter the dataset to the year 2007
3. For the country of _Mauritius_, what is the population in the year 1997?
4. How many different countries are in the dataset?


```{r gapminder-exploration-df}

```


Hint: The function `unique()` is similar to `levels()` for factors, but more general -- both can be combined with `length()`!
